---
title: "AutoEDA Packages for EDA0 (SFAN WQ I&M Data)"
author: "Emalia Partlow (CCS Biology at UCSB): emaliapartlow107@gmail.com"
date: "10/29/2020"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

This document is meant to be run chunk by chunk. Do not try to Knit.

All packages mentioned in “The Landscape of R Packages for Automated Exploratory Data Analysis” (Staniak and Biecek, 2019) except lubridate.

In "AutoEDA Packages in R for EDA0" (Partlow, 2020) and Table 1, I compared the relative strengths of autoEDA packages for EDA0 and illustrated to what extent each package performs common EDA0 techniques.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr) # Subset columns to make new df called sfanWQ (function used: select). sfanWQ long to wide format (function used: mutate). New df with only numeric variables (function used: select_if).
library(tidyverse) # sfanWQ long to wide format (function used: pivot_wider).
library(readr) # Import data as tibble (function used: read_csv)

# 10 packages highlighted in "AutoEDA Packages in R for EDA0" (Partlow, 2020) for having the most potential for the EDA0 of NPS I&M data:
library(lubridate)
library(dataMaid)
library(explore)
library(DataExplorer)
library(SmartEDA)
library(summarytools)
library(inspectdf) # Install GitHub version (devtools::install_github("alastairrushworth/inspectdf")).
library(funModeling)
library(visdat)

library(ggplot2) # Rasters with ggplot2 needs:
library(plyr)
library(reshape2)
library(data.table)
library(RColorBrewer)
library(gridExtra)


# Remaining autoEDA packages:
library(dlookr)
library(ExPanDaR) # Install GitHub version of ExPanDaR in order to get option to export report as notebook (devtools::install_github("joachim-gassen/ExPanDaR")).
library(autoEDA) # Instal from GitHub (devtools::install_github("XanderHorn/autoEDA"))
library(skimr)
library(tableone)
library(describer)
library(prettyR)
library(Hmisc)
library(arsenal)
library(xray)
```


Read in WQ dataset, manipulate:
```{r}
sfanWQoriginal <- read_csv("SFAN_WQ_PublicData_WY2006_2014.csv") # Import data as a tibble.
summary(sfanWQoriginal)
str(sfanWQoriginal)
head(sfanWQoriginal)

sfanWQ <- select(sfanWQoriginal, StationID, `Visit Start Date`, `Depth to Activity`, `Local Characteristic Name`, `Result Value/Text`, `Detection Limit`, `Lower Quantification Limit`, `Upper Quantification Limit`) # Create new data frame called sfanWQ with subseted columns for analysis.

sfanWQ2 <- sfanWQ # Copy sfanWQ.
sfanWQ2$`Result Value/Text` <- with(sfanWQ2, ifelse(`Result Value/Text` %in% "*Non-detect", `Lower Quantification Limit`, ifelse(`Result Value/Text` %in% "*Present >QL", `Upper Quantification Limit`, ifelse(`Result Value/Text` %in% "*Present <QL", `Lower Quantification Limit`, `Result Value/Text`)))) # In column `Result Value/Text', replace text *Non-detect and *Present <QL with the value in column 'Lower Quantification Limit' and replace text *Present >QL with the value in column `Upper Quantification Limit`.

sfanWQwide <- sfanWQ2 %>%
  group_by(`Local Characteristic Name`) %>%
  dplyr::mutate(rn = row_number()) %>%
  pivot_wider(names_from = `Local Characteristic Name`, values_from = `Result Value/Text`) # sfanWQ2 switched from long to wide format with a new column for each parameter (Local.Characteristic.Name). mutate() makes data type = characters not "list", new column "rn" for row #.
sfanWQwide[sfanWQwide == "*Not Reported"] <- NA # Change all *Not Reported to NAs.
sfanWQwide$Lab_or_Visit_Notes <- NULL # Removes lab/visit notes column because variables with lengthy characters are not displayed well in reports.

sfanWQwide2 <- sfanWQwide %>% dplyr::mutate(across(c(AirTemp_Glass, AirTemp_Oakton, `Ammonia as N`, `Ammonia as NH3`, Cond_YSI85, discharge, `DO%_YSI85`, DO_YSI85, E.coli_QuantiTray, E.coli_SFANLab, `Estimated Flow`, `Estimated velocity`, FecalColiform_MultiTube, `Gage Height`, H2OTemp_Oakton, H2OTemp_YSI85, `Last Precipitation`, `Nitrate as N`, `Nitrate as NO3`, pH_Oakton, pH_paper, Salinity_YSI85, SC_YSI85, SSC, `Stream Width`, `Total Kjeldahl Nitrogen`, TotalColiform_MPN, TotalColiform_QuantiTray, TotalColiform_SFANLab, TSS, Turbidity_Hach2100, `Nitrite as N`, pH_Sension1, H20Temp_sension, `H2OTemp_YSI Pro Plus`, `Cond_YSI Pro Plus`, `SC_YSI Pro Plus`, `pH_YSI Pro Plus`, `Salinity_YSI Pro Plus`, `DO%_YSI Pro Plus`, `DO_YSI Pro Plus`, DO_YSI556, `DO%_YSI556`, pH_YSI556, Salinity_YSI556, Cond_YSI556, SC_YSI556, H2OTemp_YSI556), as.numeric)) # New data frame with data types corrected (from character in sfanWQwide to numeric in sfanWQwide2).
str(sfanWQwide2)

sfanWQwide2 <- sfanWQwide2 %>% dplyr::rename(Date = "Visit Start Date", Depth = "Depth to Activity", LQL = "Lower Quantification Limit", UQL = "Upper Quantification Limit", Ammonia.as.N = "Ammonia as N", Ammonia.as.NH3 = "Ammonia as NH3", DOpercent_YSI85 = "DO%_YSI85", Estimated.Flow = "Estimated Flow", Estimated.velocity = "Estimated velocity", Gage.Height = "Gage Height", General.Notes = "General Notes", Last.Precipitation = "Last Precipitation", Nitrate.as.N = "Nitrate as N", Nitrate.as.NO3 = "Nitrate as NO3", Stream.Width = "Stream Width", Total.Kjeldahl.Nitrogen = "Total Kjeldahl Nitrogen", Water.Appearance = "Water Appearance", Nitrite.as.N = "Nitrite as N", H2OTemp_YSI.Pro.Plus = "H2OTemp_YSI Pro Plus", Cond_YSI.Pro.Plus = "Cond_YSI Pro Plus", SC_YSI.Pro.Plus = "SC_YSI Pro Plus", pH_YSI.Pro.Plus = "pH_YSI Pro Plus", Salinity_YSI.Pro.Plus = "Salinity_YSI Pro Plus", DOpercent_YSI.Pro.Plus = "DO%_YSI Pro Plus", DO_YSI.Pro.Plus = "DO_YSI Pro Plus", DOpercent_YSI556 = "DO%_YSI556") # Rename column headers to remove symbols and spaces.

sfanWQwide2_num <- select_if(sfanWQwide2, is.numeric) # New data frame that is a subset of numeric variables from sfanWQwide2.
sfanWQwide2_num$Depth <- sfanWQwide2_num$LQL <- sfanWQwide2_num$UQL <- sfanWQwide2_num$rn <- NULL # Remove columns that are not WQ parameter variables.
```


10 packages highlighted in "AutoEDA Packages in R for EDA0" for having the most potential for the EDA0 of NPS I&M data: lubridate, dataMaid, explore, DataExplorer, SmartEDA, summarytools, inspectdf, funModeling, visdat.


lubridate:
```{r}
class(sfanWQwide2$Date) # Check class of dates column.
head(sfanWQwide2$Date) # Look at how dates are formated (m/d/y).

sfanWQwide2$Date <- mdy(sfanWQwide2$Date) # Change data type from character to date.
class(sfanWQwide2$Date) # Double check class.
head(sfanWQwide2$Date) # Check format.

table(month(sfanWQwide2$Date, label = TRUE)) # Frequency table with the number of times each month appears in the Date column. Can replace month() with year(), day(), wday(). Remove label=TRUE for year, month
barplot(table(month(sfanWQwide2$Date, label = TRUE))) # Visualize frequency table as bar plot.

# To look at interval between each row in Date column use:
diff(sfanWQwide2$Date) # In days
# or
as.duration(int_diff(sfanWQwide2$Date)) # In weeks
```


dataMaid:
```{r}
makeDataReport(sfanWQwide2, output = "html", replace = TRUE, reportTitle = "SFAN WQ EDA0") # Report includes: summary of full dataset, checks performed, summary table and visualization (bar plot or histogram) for each variable, which variables are "problematic" and why they were flagged (anomaly/outlier detection). To make report with only problematic variables include onlyProblematic=TRUE. If column headers include symbols (%) or spaces a report will not be generated because of an invalid filename error. It is dIfficult to read bar plots of categorical variables with more than a few character types (the text on the x axis overlaps). Missing value percentage can be inaccurate for columns with little data.

# A report can not target specific variables, but the seperate check, summarize, and visualize functions can:

allCheckFunctions() # Can change "Check" to "Summary" or "Visual". Gives list of specific functions that can be applied to each variable.

dataMaid::summarize(sfanWQwide2$AirTemp_Glass) # Variable description, stats.
visualize(sfanWQwide2$AirTemp_Glass) # Variable distribution.
check(sfanWQwide2$AirTemp_Glass) # Details potential problem(s) with variable (includes outlier detection).
```


explore:
```{r}
explore(sfanWQwide2) # Shiny interactive explore pop up. Can specify target.

report(sfanWQwide2, output_file = "explore_report_sfanWQwide2", output_dir = "./reports") # Must specify output file and directory. Easier to visualize categorical variables than dataMaid because they are plotted along the y axis. Missing value percentage can be inaccurate for columns with little data.

report(sfanWQwide2, target = StationID, output_file = "explore_report_sfanWQwide_stationID", output_dir = "./reports") # Report all variables with binary target (e.g. StationID).
```


DataExplorer:
```{r}
create_report(sfanWQwide2, output_file = "DataExplorer_report_sfanWQwide2.html") # Report includes: dataset summary, missing data profile (missing value percentage can be inaccurate for columns with little data), data structure tree (difficult to read when it is used for a dataset with ~50+ variables), distributions (histogram or bar plot), QQPlots, correlation analysis (missing b/c insufficient complete rows), principal component analysis (missing b/c insuficient complete rows).

plot_boxplot(sfanWQwide2, by = "StationID", nrow = 1L) # Boxplot of each variable broken down by StationID. Specify 1 Row.
```


SmartEDA:
```{r}
ExpReport(sfanWQwide2, Target = "AirTemp_Glass", op_file = "SmartEDA_report_sfanWQwide2") # Report includes: summary and visualization of each vairable (Numeric variables: QQplot, density plot, scatter plot. Categorical variables: bar plots). Use ggthemes package to customize ggplots. If column headers have symbols (%, -, etc.) or spaces, a report will not be generated because of an unexpected symbol error. Unlike the other reports, correlation analysis is not skipped because of "insufficient complete rows". A bivariate scatter plot is created for each possible variable pairing, which results in multiple blank plots. It takes the longest time, compared to the other 1 line report packages, for a report to be generated (because of the scatter plots).
```


summarytools:
```{r}
summarytools::freq(sfanWQwide2) # Frequency table (counts, missing values) for valid/total data in dataset. Can create simpler tables by setting certain columns to = FALSE.

view(dfSummary(sfanWQwide2)) # Print out data frame summary in viewer with basic stats, freqs, histogram, and missing data information.

sfanWQst <- stby(data = sfanWQwide2, INDICES = sfanWQwide2$StationID, FUN = descr, stats = "common", transpose = TRUE) # Group variables by parameter. Display basic stats for each numeric variable in each group (e.g. StationID).
view(sfanWQst)
```


inspectdf:
```{r}
# Add %>% show_plot() to each function to see graphic. Functions include options for comparing 2 datasets:

inspect_na(sfanWQwide2) %>% show_plot() # Missing values for each variable. Compare between 2 datasets by adding df2 = WQwide2. Missing value percentage can be inaccurate for columns with little data.
inspect_types(sfanWQwide2) # Number of each variable type.

inspect_cat(sfanWQwide2) # Frequency of categorical levels in df.
inspect_imb(sfanWQwide2) # Summary of most common level in each categorical variable.

inspect_num(sfanWQwide2) %>% show_plot(plot_layout = c(6, 9)) # Summary/stats for numeric variable levels. Plot shows histogram of each numeric variable. plot_layout specifies the number of rows and columns in the histogram document.
inspect_cor(sfanWQwide2_num, method = "pearson") %>% show_plot() # Bivariate relationship/correlation coefficients for numeric variables. Can specify method.
```


funModeling:
```{r}
df_status(sfanWQwide2) # Summary of data frame (missing values, data type, unique values). Missing value percentage can be inaccurate for columns with little data.

funModeling::freq(sfanWQwide2) # Frequency table and bar plot for each categorical variable.

profiling_num(sfanWQwide2) # Table with multiple metrics for numeric variables.

sfanWQwide3 <- sfanWQwide2_num # Copy sfanWQwide2_num.
sfanWQwide3$Detection.Limit <- NULL # Delete Detection.Limit column in order to have a more accurate scale for the histograms (see below).
plot_num(sfanWQwide3) # 1 plot with histograms for all numeric variables. Can specify bin #. There is 1 scale for all plotted variables, so it might be difficult to visualize them all if the measurements are very different in size. Can subset the data (see above) to get a clearer graphic.

# correlation_table(data = sfanWQwide2, target = "AirTemp_Glass")# Correlation against targeted variable. Error because needs complete numeric column.

# cross_plot() and plotar() plot distribution of target vs predictor.
plotar(sfanWQwide2, target = "StationID", plot_type = "boxplot") # Boxplot of each variable broken down by StationID.
```


visdat:
```{r}
vis_dat(sfanWQwide2, warn_large_data = FALSE) # Visualize data types and NAs. Set warn_large_data = FALSE for large datasets.

vis_guess(sfanWQwide2) # Visualize value type.

vis_miss(sfanWQwide2, warn_large_data = FALSE) # Visualize missing data.

vis_compare(sfanWQwide2, sfanWQwide) # Visualize difference between 2 datasets that have the same dimensions.

vis_expect(sfanWQwide2, ~ .x == 0) # Visualize whether a value is in the dataset (e.g. 0) or where a specified condition is met.

vis_cor(sfanWQwide2_num) # Heatmap displaying correlations. Can specify what to do with NA values.
```


Rasters with ggplot2:
```{r}
# Packages needed: ggplot2, plyr. reshape2, data.table, RColorBrewer, gridExtra

ggsfanWQwide2<-sfanWQwide2 # Make copy of sfanWQwide2.

ggsfanWQwide2$DOpercent_YSI85.E <- ifelse(ggsfanWQwide2$DOpercent_YSI85 <= 50, 1, 0) # Add column with a value of 1 when DOpercent is less than 50

# Create new dataset which collapses duplicated readings in one month by taking the means of the readings,creates columns denoting when there are exceedance values, and also creates stand along columns for Year and Month:
ggsfanWQwide2$Date <- mdy(ggsfanWQwide2$Date) # Change data type from character to date.
sitestats <- ddply(ggsfanWQwide2, c("StationID", "Date"), summarize, DOpercent_YSI85=mean(DOpercent_YSI85, na.rm=TRUE))
e <- ddply(ggsfanWQwide2, c("StationID", "Date"), summarize, DOpercent_YSI85.E=sum(DOpercent_YSI85.E, na.rm=TRUE)) # Create a file that sums exccedances, so that any row NOT including any exceedances is represented with a 0, any rows containing any exceedances is represented by 1 or greater.
sitestats$Y <- year(sitestats$Date)
sitestats$M <- month(sitestats$Date)
e$StationID <- NULL
e$Date <- NULL
sitestats <- cbind(sitestats, e)
sitestats$DOpercent_YSI85.E <- ifelse(sitestats$DOpercent_YSI85.E>0, "*", NA) # In the column DOpercent_YSI85.E, change any value >0 to a * for printing on charts.


xlabels <- c("O", "N", "D", "J", "F", "M", "A", "M", "J", "J", "A", "S") # Set names for plot x axis labels (months)

siteDO <- as.data.table(sitestats) # Convert table sitestats into a data.table to enable the plots to call for specific sites.

#Call plots:
for (i in unique(ggsfanWQwide2$StationID)) { # Set up loop to cycle through all sites.
   
raster <- siteDO[StationID==i] # Create a variable that includes one site and all parameters.

name=paste("StationID", as.character(raster$StationID[1])) # Define variable name as StationID name for plot title.

p.DOpercent_YSI85 <- ggplot(raster, aes(M, Y)) + # Call a raster plot with x axis water months, y axis water years
geom_tile(aes(fill=DOpercent_YSI85)) + # Fill raster plot with temperature data.
geom_text(aes(label=DOpercent_YSI85.E)) + # Print number of exceedances in raster plot.
scale_y_continuous(breaks=unique(sitestats$Y)) + # Name y axis ticks with years.
scale_x_continuous(breaks=c(1:12), labels=xlabels) + # Name x axis ticks with months.
labs(x="Month",y="Year") + # Label x and y axes
ggtitle(paste0(name,"DOpercent_YSI85")) + # Label plot
theme(plot.title=element_text(size=10)) + # Size plot title
theme(panel.background = element_rect(fill='white', colour='grey')) + # Remove gray fill around plot.
scale_fill_gradientn(colors = (brewer.pal(11,"RdYlBu")), limits=c(min(sitestats$DOpercent_YSI85, na.rm=TRUE),
max(sitestats$DOpercent_YSI85, na.rm=TRUE))) # Set color gradient and min/max for fill pattern.

print(p.DOpercent_YSI85)
plotname <- paste(raster$StationID[1], "DOpercent_YSI85", "tiff", sep=".") # Create a file name for the plot.
ggsave(plotname, plot = last_plot(), device = "tiff", width = 6.5, height = 5, units = "in") # Save plot to working directory as a tiff.

}

p <- grid.arrange(p.DOpercent_YSI85, ncol=4) # Draw four plots in a row
```

End of 10 packages detailed in "AutoEDA Packages in R for EDA0" (Partlow, 2020)
------------------------------------------------------------------------


dlookr: 
```{r}
# eda_report(sfanWQwide2, output_format = "html", output_file = "dlookr_report_sfanWQwide2.html", output_dir = "./reports")# Can add target. Report includes: general summary of dataset, univariate analysis (normality test for numeric variables. Includes log and sqrt transformations), corelation between variables, target analysis. For large datasets Error in shapiro.test(x): sample size must be between 3 and 5000. Maybe change "collect_size =".

diagnose_report(sfanWQwide2, output_format = "html", output_file = "dlookr_diagnose_sfanWQwide2.html", output_dir = "./") # Unlike eda_report, does not result in error when used on large dataset. Includes: variable summaries/descriptions and useful outlier diagnosis of each numeric variable.
```


ExPanDaR:
```{r}
ExPanD(sfanWQwide2, df_name = "sfanWQwide2 ExPanDaR EDA Report", export_nb_option = TRUE) # Useful df_fef, var_def, abstract functions (options for attaching variable and dataset metadata). There is no summary of the overall dataset structure and missing values (although some variables show NAs in bar plots). export_nb_option set too TRUE to download as R notebook.
```


autoEDA:
```{r}
autoEDA(sfanWQwide2, removeMajorityMissing = FALSE, rotateLabels = TRUE, imputeMissing = FALSE, outputPath = "./reports") # Visualize numeric and categorical variables. Change graphics for continuous/discrete variables: plotContinuous= "boxplot", "qqplot", "density", "histogram". Change graphics for categorical variables: plotCategorical= "bar", "stackedBar", "groupedBar". Specify outputPath to print as PDF. Set removeMajorityMissing to FALSE to get graphics for variables with more than 1/2 observations missing. Set imputeMissing to FALSE so that the missing values are not imputed using the medium value. Add y= for bivariate analysis.

dataOverview(sfanWQwide2) # Tabular EDA with data type, missing values, unique values, outliers, and stats.

# Use predictivePower() for correlation analysis.
```


skimr:
```{r}
skim(sfanWQwide2) # Basic summary/description/stats of dataset and variables, basic/small histograms. Good for a quick glance into the data (similar to what is included in the beginning of data overview reports from other packages).

skim(sfanWQwide2, AirTemp_Glass, DOpercent_YSI85) # Targets and gives summaries for AirTemp_Glass and DOpercent_YSI85 parameters.
```


tableone:
```{r}
CreateTableOne(data = sfanWQwide2) # Summary of variables: frequency of non numeric, mean + SD of numeric.
```


describer:
```{r}
describer::describe(sfanWQwide2) # Basic description of variables with some simple stats (column class, mean, SD). Can target variable.
```


prettyR:
```{r}
prettyR::describe(sfanWQwide2) # Description of variables (mean, median, varience, SD, NAs, frequency of categorical variables). When specify a variable it prints out all values of that column.
```


Hmisc:
```{r}
Hmisc::describe(sfanWQwide2) # Basic description of dataset including missing values. Lists the lowest and highest values in each column (aids outlier detection).
```


arsenal:
```{r}
comparedf(sfanWQ, sfanWQwide2) # Compare different datasets. What variables do they share? For a more detailed description of the differences add summary().

summary(freqlist(~ StationID + Weather, data = sfanWQwide2), text = TRUE) # Frequency table for categorical variables. Can layer/group results so that it shows the frequency of how many times 2+ categorical variables appear together.

summary(tableby(~ StationID + AirTemp_Glass + DOpercent_YSI85, data = sfanWQwide2), text = TRUE) # Table of descriptive stats (mean, SD, range) and frequencies (including NAs) for targeted variable.

# Use modelsum() for lin reg table.
```


xray:
```{r}
anomalies(sfanWQwide2) # Searches for anomalous columns. Can change anomaly and distinct value threshholds.

distributions(sfanWQwide2) # Graphic (histogram or bar chart) of each variable. Nice + quick visualizations.

# Use timebased() to look at each variable with a time variable.
```
