---
title: "AutoEDApackages_EDA1"
author: "Nicole Hupp"
date: "10/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

library(dplyr) # Subset columns to make new df called sfanWQ (function used: select). sfanWQ long to wide format (function used: mutate). New df with only numeric variables (function used: select_if).
library(tidyverse) # sfanWQ long to wide format (function used: pivot_wider).
library(readr) # Import data as tibble (function used: read_csv)

# 9 packages highlighted in "AutoEDA Packages in RStudio for EDA0" (Partlow, 2020) for having the most potential for the EDA0 of NPS I&M data:

library(lubridate)
library(dataMaid)
library(explore)
library(DataExplorer)
library(SmartEDA)
library(summarytools)
library(inspectdf) # Install GitHub version (devtools::install_github("alastairrushworth/inspectdf")).
library(funModeling)
library(visdat)

# Remaining autoEDA packages:

library(dlookr)
library(ExPanDaR) # Install GitHub version of ExPanDaR in order to get option to export report as notebook (devtools::install_github("joachim-gassen/ExPanDaR")).
library(autoEDA) # Instal from GitHub (devtools::install_github("XanderHorn/autoEDA"))
library(skimr)
library(tableone)
library(describer)
library(prettyR)
library(Hmisc)
library(arsenal)
library(xray)
```

Read in MOJN IU dataset, manipulate:
```{r}